---
output:
  html_document: default
  pdf_document: default
---

<!-- # Filename: Oregon_Crime_Report.rmd -->

<!-- # Author: @russl_corey -->

<!-- # Date: oct 10, 2022 -->

<!-- # This program is free software: you can redistribute it and/or modify it under  -->

<!-- # the terms of the GNU General Public License as published by the Free Software  -->

<!-- # Foundation, either version 3 of the License, or (at your option) any later  -->

<!-- # version. -->

<!-- #  -->

<!-- # This program is distributed in the hope that it will be useful, but WITHOUT ANY  -->

<!-- # WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A  -->

<!-- # PARTICULAR PURPOSE. See the GNU General Public License for more details. -->

<!-- # You should have received a copy of the GNU General Public License along with  -->

<!-- # this program. If not, see <https://www.gnu.org/licenses/>.  -->

---
title: "Oregon Open Crime Data Analysis"
author: Russell_Corey
date: October, 19 2022
output: html_document
---

\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
```

```{r load crime data, echo=FALSE, message=FALSE, warning=FALSE}
source("Oregon_Crime_Offenses.r")

setwd('/home/russell/Dropbox/DataAnalysis/Oregon_Crime_Report')

# define study counties
study_counties = c('lane', 'multnomah', 'washington', 'baker', 'clackamas')
```

# Intro

We will explore crime reports in Oregon. We will look at the number of reported offenses as aggregated by both month and county along with a close look at violent offenses in particular. Rates of offense aggregated along counties will be normalized using 2020 Census data. Source code will be posted on [Github](https://github.com/russl-corey/Oregon_Crime)

# Data

The two data sources used are the "Uniform Crime Reporting" Offenses data and the Oregon 2020 Census data from Portland University.

#### Crime Data

The crime reporting data set is accissable directly from

<https://www.oregon.gov/osp/Pages/Uniform-Crime-Reporting-Data.aspx>

#### Population Data

The Oregon census data can be found at the website

<https://www.pdx.edu/population-research/census-data-oregon>

with the direct link being

<https://drive.google.com/uc?export=download&id=1JrrmYiQUBPux8nnJ88epAAk9U5rbRDBD>

\newpage

# Data Exploration and Validation

Overview of number of records collected per month and the total number of distinct offenses for the data set.

```{r sample offenses, echo=FALSE, message=FALSE}
#library(kableExtra)

all_offenses %>%
  tail() %>%
  select(-c('date', 'county', 'NIBRS Crime Against')) %>%
  kable(caption = 'Offenses Reports') #%>%
  #kable_styling( font_size = 5)

```

Each record has a count of distinct offenses. We can compare the number of reports per month and the total number of distinct offenses.

```{r define helper functions, echo=FALSE, message=FALSE}
all_offenses_m_count <- tally_off_month(all_offenses)
p <- ggplot() +
  labs(y='Number per Month', title="Open Data Offenses Overview") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 month") +
  geom_line(data=all_offenses_m_count, aes(x=month, y=`Total Distinct Offenses`, color='Number Distinct Offenses')) +
  geom_line(data=all_offenses_m_count, aes(x=month, y=`Total Records`, color='Number of Records'))
p

all_offenses_m_count %>%
  select('month', 'Total Records', 'Total Distinct Offenses') %>%
  tail() %>%
  kable(caption = 'Recent Offenses')

```

There is a shape decline in the last month. We will ignore that month, assuming that it is due to a lag in reporting and assimilating.

## Outlier Overview

```{r total offenses boxplot, echo=FALSE, message=FALSE}

boxplot(all_offenses_m_count$`Total Distinct Offenses`, ylab='Distinct Offenses', main='Potential Outliers: "Total Distinct Offenses"')

out <- boxplot.stats(all_offenses_m_count$`Total Distinct Offenses`)$out
out_ind <- which(all_offenses_m_count$`Total Distinct Offenses` %in% c(out))

all_offenses_m_count[out_ind,] %>%
  select('month', 'Total Records', 'Total Distinct Offenses') %>%
  mutate('z_score' = round((`Total Distinct Offenses` - mean(all_offenses_m_count$`Total Distinct Offenses`)) / sd(all_offenses_m_count$`Total Distinct Offenses`), digits=2) ) %>%
  kable(caption='Potential Outliers: "Total Distinct Offenses"')

```

```{r total records boxplot, echo=FALSE, message=FALSE}

boxplot(all_offenses_m_count$`Total Records`, ylab='Total Records', main='Potential Outliers: "Total Records"')

out <- boxplot.stats(all_offenses_m_count$`Total Records`)$out
out_ind <- which(all_offenses_m_count$`Total Records` %in% c(out))

all_offenses_m_count[out_ind,] %>%
  select('month', 'Total Records', 'Total Distinct Offenses') %>%
  mutate('z_score' = round((`Total Records` - mean(all_offenses_m_count$`Total Records`)) / sd(all_offenses_m_count$`Total Records`), digits=2) ) %>%
  kable(caption='Potential Outliers: "Total Records"')

```

We will then be excluding month 33, '2022-09-01'.

# Population

For per ca-pita calculations we need to know the population broken down by counties. This data is in the Census data set. The data gives us the population in 2010 and 2020 along with the percentage change.

```{r plot, echo=FALSE, results = 'asis', warning=FALSE, }


# Show table of populations in study counties
county_pop %>%
  subset(county_pop$county %in% study_counties) %>%
  select('county', '2010', '2020', 'percent_change', 'pop_rate_year') %>%
  kable(caption = "County Population")

```

# Violent Offenses

Violent offenses are of particular concern so let's look at those offenses. To determine which offenses are violent we use the FBI's definition of violent crimes to categorize the 'NIBRS Report Title' column in the data set.

```{r violent titles, echo= FALSE, results = 'asis', warning=FALSE}

kable(violent_titles, caption = 'Violent Offense Titles')
```

We can look at the break down of these violent offenses for the whole state.

```{r violent offenses overview by type, echo= FALSE, results = 'asis', warning=FALSE}

# tally offenses 
violent_m_count <- tally_vo_month(offenses)

# Line plot monthly violent offenses
p <- ggplot() +
  labs(y='Number per Month', title=paste("Violent Offenses in Oregon", start_month, 'to',  end_month)) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 month") +
  geom_line(data=violent_m_count, aes(x=month, y=`All Violent Offenses`, color='All Violence')) +
  geom_line(data=violent_m_count, aes(x=month, y=`Aggravated Assault`, color="Assault")) +
  geom_line(data=violent_m_count, aes(x=month, y=`Robbery`, color="Robbery")) +
  geom_line(data=violent_m_count, aes(x=month, y=`Forcible Rape`, color="Rape")) +
  geom_line(data=violent_m_count, aes(x=month, y=`Willful Murder`, color="Murder")) 

p

violent_m_count %>%
  tail() %>%
  select(-c('m', 'v_off', 'Total Distinct Offenses')) %>%
  kable(caption = paste('Monthly Violent Offenses', start_month, 'to', end_month))#,
     #   format='latex') # %>%
  #kable_styling( font_size = 6, full_width = TRUE)

```

We can also compare the number of violent offenses verses non-violent offenses for the state over the study period.

```{r violent offenses overview comparison, echo= FALSE, results = 'asis', warning=FALSE}

# Line plot violent offenses vs nonviolent offenses
  p <- ggplot() +
    labs(y='Number per Month', title=paste("Comparison of Offenses in Oregon", start_month, 'to', end_month)) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "6 month") +
    geom_line(data=violent_m_count, aes(x=month, y=`All Violent Offenses`, color='Violent Offenses')) +
    geom_line(data=violent_m_count, aes(x=month, y=`Nonviolent Offenses`, color="Nonviolent Offenses")) 
  p  
```

Next we can look at the total number of distinct violent offenses by county.

```{r violent offenses by county ,  echo= FALSE, results = 'asis', warning=FALSE}

# Combined line plot total violent offenses in select counties
lane = tally_vo_county_month( 'lane')
washington = tally_vo_county_month('washington')
multnomah = tally_vo_county_month('multnomah')
washington = tally_vo_county_month('washington')
linn = tally_vo_county_month('linn')

 p <- ggplot() +
    labs(y='Number per Month', title="Total Violent Offenses in Select Counties") +
    scale_x_date(date_labels = "%b %Y", date_breaks = "6 month") +
    geom_line(data=lane, aes(x=month, y=`All Violent Offenses`, color='Lane County')) +
    geom_line(data=washington, aes(x=month, y=`All Violent Offenses`, color='Washington County')) +
   geom_line(data=linn, aes(x=month, y=`All Violent Offenses`, color='Linn County')) +
    geom_line(data=multnomah, aes(x=month, y=`All Violent Offenses`, color='Multnomah County')) 
  p  
```

Recall from the population table that these counties vary greatly by population. So, using the population data we can plot the above graph adjusted to the county's population.

# Violent Offenses by County

We can look at the aggregates of violent offenses by county.

```{r violent offenses by county per capita table2,  echo= FALSE, results = 'asis', warning=FALSE}
vo_county_sums %>%
#  subset(`county` %in% study_counties) %>%
  kable( caption = 'County Total Violent Offenses', digits=2)

barplot(vo_county_sums$offenses,
  main = "Total Violent Offenses Over Study Period", 
  ylab = "Offenses",
  names.arg = vo_county_sums$county,
  #col = "blue",
  density = 40,
  las = 2,
  cex.names = 0.7,
  #horiz = TRUE
  )
```

From the table it appears that the rate of violent offenses is far higher in Multnomah county than others.

```{r violent offenses by county per capita table3,  echo= FALSE, results = 'asis', warning=FALSE}

merged_data <- vo_county_sums %>% merge(county_pop, by='county', all=FALSE) 

merged_data[order(merged_data$vo_percap, decreasing = TRUE),] %>%
  select('county', 'offenses', '2020', 'vo_percap') %>%
 # head() %>%
  kable( caption = 'Counties with Highest Per Capita Violent Offenses ', digits=2)

```



```{r violent offenses by county per capita map,  echo= FALSE, results = 'asis', warning=FALSE}
#library(maptools)
library(sf)
library(ggplot2)

area <- read_sf("data/cb_2018_us_county_500k/cb_2018_us_county_500k.shp")

# subset only oregon counties
area <- subset(area, area$STATEFP == 41) 

# rename counties to lower case
area$county <- tolower(area$NAME)
area<- merge(x=area, y=vo_county_sums, by='county', all=FALSE)

# define IDs for the polygons
area$id = rownames(area)

#area.map <- fortify(area, region = 'id')

ggplot() +
  geom_sf(data=area, aes(fill =vo_percap)) +
  scale_fill_gradient(low='white', high='red', na.value= NA) +
  labs(title='Per Capita Violent Offenses', fill='Violent Offense Per Capita') +
   theme_void() 
```

# Corrlation

```{r violent offenses by county corr matrix,  echo= FALSE, results = 'asis', warning=FALSE}
library("ggcorrplot")       

county_violent_month %>%
  select(-c('month')) %>%
  cor() %>%
  ggcorrplot(show.diag = FALSE, type = 'lower') +  
  ggplot2::theme(axis.text.y=ggplot2::element_text(size=10),
                 axis.text.x=ggplot2::element_text(size=8),
                 legend.text=ggplot2::element_text(size=10))
  #scale_x_discrete(guide = guide_axis(n.dodge=2))+
 # scale_y_discrete(guide = guide_axis(n.dodge=2)) +
  #geom_tile(fill="white") +
#  geom_tile(height=1.2, width=1.2) 
  #theme(text = element_text(size = 12))
```

### Total monthly violent offenses by county

```{r violent offenses by county monthly ,  echo= FALSE, results = 'asis', warning=FALSE}

# Combined line plot total violent offenses in select counties
cnty_violent_cor %>%
  filter(`pearson_cor` > 0.32) %>%
  select('pearson_cor') %>%
  kable(row.names = NA, caption='Counties with positiv correlation >0.32 ', digits=2)

cnty_violent_cor %>%
  filter(`pearson_cor` < -0.32) %>%
  select('pearson_cor') %>%
  kable(caption='Counties with negative correlation < -0.32 ', digits=2)
```

# Linear Regression

```{r violent offenses by county plots,  echo= FALSE, results = 'asis', warning=FALSE}
    

#for(countyname in study_counties){
countyname <- 'oregon'
  print(
  county_violent_month %>%
    select('x', all_of(countyname)) %>%
    rename('month' = 'x') %>%
    rename('v_off' = all_of(countyname)) %>%
   # rename('m' = x) %>%
    ggplot( aes(month, v_off)) +
    labs(title = paste(countyname,
                      'pearson corr:', lm_summary$pearson_cor[lm_summary$county == countyname],
                      'vo_rate_year:', lm_summary$vo_rate_year[lm_summary$county == countyname])) +
    geom_point() +
    geom_smooth(method='lm')
  )
#}

#data

```

```{r violent offeneses month lin regg,  echo= FALSE, results = 'asis', warning=FALSE}

lm_summary[order(-lm_summary$pearson_cor),] %>%
  select('county', 'pearson_cor', '2020', 'vo_rate_year', 'pop_rate_year', 'adjusted_rate') %>%
  head() %>%
  kable(caption='pearson correlation and linear reg rate', digits=2)

lm_summary[order(-lm_summary$pearson_cor),] %>%
  select('county','pearson_cor', '2020',  'vo_rate_year', 'pop_rate_year', 'adjusted_rate') %>%
  tail() %>%
  kable(caption='pearson correlation and linear reg rate', digits =2)
```

\newpage

# Sources

#### Oregon Crime Data

"Oregon State Police: Uniform Crime Reporting"

    https://www.oregon.gov/osp/Pages/Uniform-Crime-Reporting-Data.aspx

#### Oregon Census Data

"Census Data for Oregon" Population Research Center

    https://www.pdx.edu/population-research/census-data-oregon 

Direct link for data set:

    https://drive.google.com/uc?export=download&id=1JrrmYiQUBPux8nnJ88epAAk9U5rbRDBD

##### FBI

"FBI-Violent Crime" FBI: UCR

    https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/violent-crime

#### Oregon Spatial Data Library

"Oregon Counties - 2015"

    https://spatialdata.oregonexplorer.info/geoportal/details;id=361c06fee9de4e24a72e280fb386a771

\newpage

# APPENDIX A

Appendix description: Raw data tables.

# A1 Total Violent Offenses Per Capita per County

Total counts for study period:

```{r violent offenses by county per capita table,  echo= FALSE, results = 'asis', warning=FALSE}

kable(vo_county_sums, caption = 'County Total Violent Offenses')
```

\newpage

# A2 Population

```{r Census data, echo=FALSE, results = 'asis'}

# Show table of populations in study counties
kable(select(county_pop, 'county', '2010', '2020', 'percent_change'), caption = "County Population 2010 and 2020")

```

\newpage

# A4 Sums of offenses by type by date

```{r offenses by type by month, echo= FALSE, results = 'asis'}

# Show table of violent crimes committed over study time
off_per_month %>% 
  select(c(1:13)) %>%
  head() %>%
  kable(caption = "Aggregate Monthly Counts") #%>%
  #kable_styling( font_size = 5)

```

\newpage

# A5 Correlation Matrix of Total Monthly Violent Offenses

```{r appendix a5, echo= FALSE, results = 'asis'}

# cnty_violent_cor %>%
#   kable(caption = 'Correlation Matrix County Monthly Violent Offenses')

lm_summary %>%
  select('county','pearson_cor', '2020',  'vo_rate_year', 'pop_rate_year', 'adjusted_rate') %>%
  kable(caption='pearson correlation and linear reg rate')
```

\newpage

# A6 Violent offenses per county

```{r appendix a6,  echo= FALSE, results = 'asis', warning=FALSE}
for(countyname in county_list){
  print(
  county_violent_month %>%
    select('x', countyname) %>%
    rename('y' = countyname) %>%
    ggplot( aes(x, y)) +
    labs(title = paste(countyname, lm_summary$pearson_cor[lm_summary$county == countyname],
         "yearly change:", lm_summary$vo_change_year[lm_summary$county == countyname], 
         "yearly pop change",lm_summary$pop_rate_year[lm_summary$county == countyname]))  +
    geom_point() +
    geom_smooth(method='lm')
  )
}

```

\newpage

# Appendix B

source code available at <https://github.com/russl-corey/Oregon_Crime>

This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see [\<https://www.gnu.org/licenses/\>](https://www.gnu.org/licenses/).
